Synthesising Termination and Non-Termination Proofs for Bit-Vector Programs

### 2014-07-07 papers/termination/paper.tex
nudges

\section{Satisfiability of a Formula in the Synthesis Fragment} \label{sec:synthesis}

In this section we present a solver for the synthesis fragment:
 \[
  \exists P,~ x . ~\forall~ y . \sigma(P, x, y)
 \]
%
where $P$ ranges over sets, while $x$ and $y$ range over ground terms (as
mentioned before, we drop the vector notation and write $x$ for $\vec{x}$).

@@

The logic we use to express the synthesis formula, i.e.~the\\
\mbox{(non-)termination} specifications, is a subset of C that we call
\newC.  The characteristic property of a \newC program is that safety can be
decided for it using a single query to a Bounded Model Checker.  A~\newC
program is just a C program with the following syntactic restrictions:
 all loops in the program must have a constant bound;
 all recursion in the program must be limited to a constant depth;
 all arrays must be statically allocated (i.e. not using \texttt{malloc}),
 and be of constant size.
Additionally, \newC programs may use nondeterministic values, assumptions
and arbitrary-width types.

@@

Concretely, {\sc synth} is implemented as shown in
Figure~\ref{fig:synth-dfd}.  We start by taking the termination
specification and generating a C program which takes as input a candidate
$(P, x_0)$ and asserts that the candidate fails to meet the specification on
at least on of the elements of {\sc inputs}.  So this program is unsafe iff
there is some candidate $(P, x_0)$ satisfying the specification for all the
elements of {\sc inputs}.  Finding a new candidate solution is therefore
reduced to model checking this C program, which by construction is loop
free.  The model checking is done using a combination of bounded model
checking, explicit state enumeration and genetic programming (GP).  The
latter two techniques involve combining the previously generated C program,
along with code that searches for candidate solutions.  Finally, if any of
the model checkers finds a candidate solution it returns it.  The form of
this candidate solution is yet another program, this time written in the
proof language $\mathcal{L}$.

@@

\begin{theorem}
 \label{thm:synth-semicomplete}
 If the existential first-order theory used to express the specification $\sigma$ is decidable,
 algorithm~\ref{alg:cegis} is semi-complete -- if a program $P \models \sigma$
 exists then algorithm~\ref{alg:cegis} will terminate.  However, if no program
 satisfies the specification, the algorithm may not terminate.  Furthermore if the domain of inputs $X$
 is finite, the algorithm is guaranteed to terminate.
\end{theorem}

\begin{proof}
 Since the {\sc ExplicitSearch} routine enumerates all programs (as can be seen by induction on
 the program length $l$), it will eventually enumerate a program that meets the specification
 in whatever set of inputs are currently being tracked.  Since the first-order theory is
 decidable, the query in {\sc Verif} will succeed for this program, causing the algorithm to terminate.
 The set of correct programs is therefore recursively enumerable and algorithm~\ref{alg:cegis}
 enumerates this set, so it is semi-complete.

@@

To evaluate our algorithm, we implemented two tools: one to generate a
termination specification from a C program, and one to synthesise a proof
from the specification.  We ran the resulting termination prover on 47
benchmarks taken from the literature and SV-COMP'15~\cite{svcomp15}.  To
provide a comparison point, we also ran {\sc ARMC}~\cite{armc-website} on
the same benchmarks.  Each tool was given a time limit of $180\,s$, and was
run on an unloaded 8-core 3.07\,GHz Xeon X5667 with 50\,GB of RAM.  The
results of these experiments are shown in Figure~\ref{fig:experiments}.

@@

Given the lack of research in handling not-linear programs, as well as
programs over bit vectors and floats, our work focused on covering these
areas.  One of the obvious conclusions that can be reached from
Figure~\ref{fig:handletable} is that most methods tend to specialise on a
certain aspect of termination proving that they can solve efficiently. 
Conversely to this view, we aim for generality, as we do not restrict the
form of the synthesised ranking functions, nor the form of the input
programs.

@@

Proving program termination implies the simultaneous search for a
termination argument and a supporting invariant.  Brock\-schmidt et
al.~share the same representation of the state of the termination proof
between the safety prover and the ranking function synthesis
tool~\cite{DBLP:conf/cav/BrockschmidtCF13}. 
In~\cite{DBLP:conf/cav/BradleyMS05}, Bradley et al.~combine the generation
of ranking functions with the generation of invariants to form a single
constraint solving problem such that the necessary supporting invariants for
the ranking function are discovered on demand.  In our setting, both the
ranking function and the supporting invariant are iteratively constructed in
the same refinement loop.

While program termination has been extensively studied, much less research
has been conducted in the area of proving non-termination.  Gupta et
al.~dynamically enumerate lasso-shaped candidate paths for counterexamples,
and then statically prove their
feasibility~\cite{DBLP:conf/popl/GuptaHMRX08}. 
In~\cite{DBLP:conf/tacas/ChenCFNO14}, Chen et al.~prove nontermination via
reduction to safety proving.  Their iterative algorithm uses counterexamples
to a fixed safety property to refine an under-approximation of a program.

>>>>

\section{Satisfiability of a Synthesis Fragment Formula} \label{sec:synthesis}

In this section we present a solver for the synthesis fragment:
\[
  \exists P,~ x . ~\forall~ y . \sigma(P, x, y)
\]
%
where $P$ ranges over sets, while $x$ and $y$ range over ground terms (as
mentioned, we drop the vector notation and write $x$ for $\vec{x}$).

@@

The logic we use to express the synthesis formula, i.e.~the
\mbox(non-)ter\-mi\-nation specifications, is a subset of C that we call
\newC.  The characteristic property of a \newC program is that safety can be
decided for it using a single query to a Bounded Model Checker.  A~\newC
program is just a C program with the following syntactic restrictions:
 all loops in the program must have a constant bound;
 all recursion in the program must be limited to a constant depth;
 all arrays must be statically allocated (i.e. not using \texttt{malloc}),
 and be of constant size.
Additionally, \newC programs may use nondeterministic values, assumptions
and arbitrary-width types.

@@

Concretely, {\sc synth} is implemented as shown in
Figure~\ref{fig:synth-dfd}.  We start by taking the termination
specification and generating a C program which takes as input a candidate
$(P, x_0)$ and asserts that the candidate fails to meet the specification on
at least on of the elements of {\sc inputs}.  So this program is unsafe iff
there is some candidate $(P, x_0)$ satisfying the specification for all the
elements of {\sc inputs}.  Finding a new candidate solution is therefore
reduced to model checking this C program, which by construction is loop
free.  The model checking is done using a combination of bounded model
checking, explicit state enumeration and genetic programming (GP).  The
latter two techniques involve combining the previously generated C program,
along with code that searches for candidate solutions.  Finally, if any of
the model checkers finds a candidate solution, it returns it.  The form of
this candidate solution is yet another program, this time written in the
proof language $\mathcal{L}$.

@@

\begin{theorem}
 \label{thm:synth-semicomplete}
 If the existential first-order theory used to express the specification $\sigma$ is decidable,
 Algorithm~\ref{alg:cegis} is semi-complete -- if a program $P \models \sigma$
 exists then Algorithm~\ref{alg:cegis} will terminate.  However, if no program
 satisfies the specification, the algorithm may not terminate.  Furthermore if the domain of inputs $X$
 is finite, the algorithm is guaranteed to terminate.
\end{theorem}

\begin{proof}
 Since the {\sc ExplicitSearch} routine enumerates all programs (as can be seen by induction on
 the program length $l$), it will eventually enumerate a program that meets the specification
 in whatever set of inputs are currently being tracked.  Since the first-order theory is
 decidable, the query in {\sc Verif} will succeed for this program, causing the algorithm to terminate.
 The set of correct programs is therefore recursively enumerable and Algorithm~\ref{alg:cegis}
 enumerates this set, so it is semi-complete.

@@

To evaluate our algorithm, we implemented two tools: one to generate a
termination specification from a C program, and one to synthesise a proof
from the specification.  We ran the resulting termination prover on 47
benchmarks taken from the literature and SV-COMP'15~\cite{svcomp15}.  To
provide a comparison point, we also ran {\sc ARMC}~\cite{armc-website} on
the same benchmarks.  Each tool was given a time limit of $180\,s$, and was
run on an unloaded 8-core 3.07\,GHz Xeon X5667 with 50\,GB of RAM.  The
results of these experiments are given in Figure~\ref{fig:experiments}.

@@

Given the lack of research on termination of not-linear programs, as well as
programs over bit vectors and floats, our work focused on covering these
areas.  One of the obvious conclusions that can be reached from
Figure~\ref{fig:handletable} is that most methods tend to specialise on a
certain aspect of termination proving that they can solve efficiently. 
Conversely to this view, we aim for generality, as we do not restrict the
form of the synthesised ranking functions, nor the form of the input
programs.

@@

Proving program termination implies the simultaneous search for a
termination argument and a supporting invariant.  Brock\-schmidt et
al.~share the same representation of the state of the termination proof
between the safety prover and the ranking function synthesis
tool~\cite{DBLP:conf/cav/BrockschmidtCF13}.  Bradley et al.~combine the
generation of ranking functions with the generation of invariants to form a
single constraint solving problem such that the necessary supporting
invariants for the ranking function are discovered on
demand~\cite{DBLP:conf/cav/BradleyMS05}.  In our setting, both the ranking
function and the supporting invariant are iteratively constructed in the
same refinement loop.

While program termination has been extensively studied, much less research
has been conducted in the area of proving non-termination.  Gupta et
al.~dynamically enumerate lasso-shaped candidate paths for counterexamples,
and then statically prove their
feasibility~\cite{DBLP:conf/popl/GuptaHMRX08}.  Chen et al.~prove
nontermination via reduction to safety
proving~\cite{DBLP:conf/tacas/ChenCFNO14}.  Their iterative algorithm uses
counterexamples to a fixed safety property to refine an under-approximation
of a program.
